Follow-up Samples

### 2016-12-15 submission/manuscript_outline_20161024.tex
update manuscript with new fig 4 and p-values

Colorectal cancer (CRC) continues to be a worldwide health problem with
early detection being used as a key component in mitigating deaths due
to the disease. Previous research suggests a link between stool
bacterial microbiome and CRC. In this study, we used a model based on
the microbiome, demographics, and prior medical history to classify
individuals as having a lesion (i.e.~adenoma or carcinoma). We then used
this model to characterize the change in the gut microbiota before and
after surgery. The overall objective was to investigate the changes in
the microbiome after surgery in patients with adenomas or carcinomas.
This model was tested on a 66 person group that included samples before
and after treatment to allow for the assessment of how the model adjusts
risk after treatment. The model used for prediction had an AUC of 0.813.
For the follow up samples our Random Forest model significantly
decreased the positive probability of lesion compared to the initial
samples for both adenoma (P-value = 7.95e-08) and carcinoma (P-value =
7.95e-08). Our model predicted that 36.4\% of the 67-person cohort had
normal colons and 63.6\% had a lesion. Some OTUs that changed the most
before and after treatment included OTUs that were affiliated with
members of Blautia, Clostridium\_XlVa and Escherichia/Shigella. Our
model suggests that treatment does significantly reduce the probability
of having a colonic lesion. Further surveillance of these individuals
will enable us to determine whether models such as the one we present
here can also be used to predict recurrence of colorectal cancer.

@@

\textbf{\emph{Outcome of Model Training}} The range of the AUC for model
training ranged from a minimum of 0.795 to a maximum of 0.854 with the
middle of all 100 runs having an AUC of 0.819. Interestingly, the worst
AUC model from training performed the best on it's respective 80/20
split test data {[}Figure 3{]}. In fact the 80/20 test performance
showed that the AUC for the middle model chosen was the most stable
(best training model test set AUC = 0.682, middle training model test
set AUC = 0.815, worse training model test set AUC = 0.931). That is to
say it had the smallest change in AUC in comparison to the minimum and
maximum AUC trained models. The middle model was close to the full
training data AUC 0.813. There was no significant difference between the
AUC of the best and middle training models (P-value = 0.419). There was
also no difference in the middle model versus worse (P-value = 0.178) or
full data model (P-value = 1). The two comparisons with a significant
difference were between the worse training model and the best training
model (P-value = 7.94e-04) and the worse training model and full
training model (P-value = 3.52e-03).

\textbf{\emph{Most Important Variables to the Model}} Overall, there
were a total of 32 variables identified as being present in more than
50\% of the training models {[}Table S2{]}. The top 5 most important
bacterial OTUs were Bacteria (Otu000013), Escherichia/Shigella
(Otu000018), Bacteria (Otu000020), Ruminococcus (Otu000017), and
Porphyromonas (Otu000153). These 5 OTUs were present in at least 99 out
of the total 100 different 80/20 runs.

\textbf{\emph{Surgical Removal of an Adenoma or Carcinoma Results in a
Decrease in Positive Probability Prediction}} A total of 1 sample was
omitted from the original 67 test sample set since it was missing a
complete set of follow up data. This left a total of 66 samples for test
predictions. After multiple comparison correction there was a
significant overall decrease in positive probability of a carcinoma and
adenoma (P-value = 1.14e-11) {[}Figure 4{]}. This decrease was
significant for both adenoma (P-value = 7.95e-08 {[}Figure 4a{]} and
carcinoma (P-value = 7.95e-08) {[}Figure 4b{]} alone. This decrease in
probability of lesion also held specifically for those with screen
relative neoplasias (SRN) (P-value = 5.35e-04). A total of 66 or 100\%
of all samples were correctly predicted to have a lesion. Although there
was a decrease in positive probability only 24 of the total 66
individuals were classified as adenoma or carcinoma free on follow up
(successful classification of 37.9\%). There was no significant
difference between the predictions and actual diagnosis for the initial
samples in the 67-sample cohort test set. However, the predictions were
significantly discordent with the diagnosis for the follow up samples
(P-value = 4.19e-10). Although there were discordent results the
respective sensitivity for the initial group was 100\% and for follow up
was 100\%, respectively.

There was 1 individual who still clearly had CRC on follow up as well as
5 individuals whose status on follow up was unknown. Although the 1
individual had a decrease in positive probability their follow up sample
was still higher than the cutoff threshold of 0.5 (positive probability
= 0.927). Interestingly, 2 individuals who were unknown on follow up
still were over the threshold cutoff of 0.5 even though, like the 1
individual with clear CRC on follow up, the probability of an adenoma or
carcinoma decreased {[}Table S3{]}.

The follow up positive probabilities were not affected by either
chemotherapy treatment (uncorrected P-value = 0.919) or radiation
therapy (uncorrected P-value = 1). There was also no difference in the
amount of change in the positive probability based on whether
individuals received chemotherapy (unccorected P-value = 0.578) or
radiation therapy (uncorrected P-value = 0.904).

@@

\textbf{Figure 3: Graph of the Receiver Operating Characteristic Curve
on Test Set Performance of the Best, Middle, Worse, and Full Training
Models.} For each of the 100 training cohort sets used had 392
individuals and the testing cohort sets had 13 individuals. The AUC on
the test sets for the best, middle, and worse models from training were
0.682, 0.815, and 0.931, respectively. cvAUC is the 20 times repeated
10-fold cross-validated AUC from training.

>>>>

Colorectal cancer (CRC) continues to be a worldwide health problem with
early detection being used as a key component in mitigating deaths due
to the disease. Previous research suggests a link between stool
bacterial microbiome and CRC. In this study, we used a model based on
the microbiome, demographics, and prior medical history to classify
individuals as having a lesion (i.e.~adenoma or carcinoma). We then used
this model to characterize the change in the gut microbiota before and
after surgery. The overall objective was to investigate the changes in
the microbiome after surgery in patients with adenomas or carcinomas.
This model was tested on a 66 person group that included samples before
and after treatment to allow for the assessment of how the model adjusts
risk after treatment. The model used for prediction had an AUC of 0.763.
For the follow up samples our Random Forest model significantly
decreased the positive probability of lesion compared to the initial
samples for both adenoma (P-value = 3.64e-11) and carcinoma (P-value =
7.95e-08). Our model predicted that 36.4\% of the 67-person cohort had
normal colons and 63.6\% had a lesion. Some OTUs that changed the most
before and after treatment included OTUs that were affiliated with
members of Blautia, Clostridium\_XlVa and Escherichia/Shigella. Our
model suggests that treatment does significantly reduce the probability
of having a colonic lesion. Further surveillance of these individuals
will enable us to determine whether models such as the one we present
here can also be used to predict recurrence of colorectal cancer.

@@

\textbf{\emph{Outcome of Model Training}} The range of the AUC for model
training ranged from a minimum of 0.723 to a maximum of 0.795 with the
middle of all 100 runs having an AUC of 0.761. Interestingly, the worst
AUC model from training performed the best on it's respective 80/20
split test data {[}Figure 3{]}. In fact the 80/20 test performance
showed that the AUC for the middle model chosen was the most stable
(best training model test set AUC = 0.646, middle training model test
set AUC = 0.744, worse training model test set AUC = 0.904). That is to
say it had the smallest change in AUC in comparison to the minimum and
maximum AUC trained models. The middle model was close to the full
training data AUC 0.763. There was no significant difference between the
AUC of the best and middle training models (P-value = 1). There was also
no difference in the middle model versus worse (P-value = 0.0431) or
full data model (P-value = 1). The two comparisons with a significant
difference were between the worse training model and the best training
model (P-value = 6.83e-04) and the worse training model and full
training model (P-value = 1.2e-03).

\textbf{\emph{Most Important Variables to the Model}} Overall, there
were a total of 37 variables identified as being present in more than
50\% of the training models {[}Table S2{]}. The top 5 most important
bacterial OTUs were Bacteria (Otu000013), Escherichia/Shigella
(Otu000018), Bacteria (Otu000020), Ruminococcus (Otu000017), and
Porphyromonas (Otu000153). These 5 OTUs were present in at least 90 out
of the total 100 different 80/20 runs.

\textbf{\emph{Surgical Removal of an Adenoma or Carcinoma Results in a
Decrease in Positive Probability Prediction}} A total of 1 sample was
omitted from the original 67 test sample set since it was missing a
complete set of follow up data. This left a total of 66 samples for test
predictions. After multiple comparison correction there was a
significant overall decrease in positive probability of a carcinoma and
adenoma (P-value = 1.11e-11) {[}Figure 4{]}. This decrease was
significant for both adenoma (P-value = 3.64e-11 {[}Figure 4a{]} and
carcinoma (P-value = 7.95e-08) {[}Figure 4b{]} alone. This decrease in
probability of lesion also held specifically for those with screen
relative neoplasias (SRN) (P-value = 7.63e-06). A total of 66 or 100\%
of all samples were correctly predicted to have a lesion. Although there
was a decrease in positive probability only 24 of the total 66
individuals were classified as adenoma or carcinoma free on follow up
(successful classification of 37.9\%). There was no significant
difference between the predictions and actual diagnosis for the initial
samples in the 67-sample cohort test set. However, the predictions were
significantly discordent with the diagnosis for the follow up samples
(P-value = 4.19e-10). Although there were discordent results the
respective sensitivity for the initial group was 100\% and for follow up
was 100\%, respectively.

There was 1 individual who still clearly had CRC on follow up as well as
5 individuals whose status on follow up was unknown. Although the 1
individual had a decrease in positive probability their follow up sample
was still higher than the cutoff threshold of 0.5 (positive probability
= 0.903). Interestingly, 1 individuals who were unknown on follow up
still were over the threshold cutoff of 0.5 even though, like the 1
individual with clear CRC on follow up, the probability of an adenoma or
carcinoma decreased {[}Table S3{]}.

The follow up positive probabilities were not affected by either
chemotherapy treatment (uncorrected P-value = 0.621) or radiation
therapy (uncorrected P-value = 0.255). There was also no difference in
the amount of change in the positive probability based on whether
individuals received chemotherapy (unccorected P-value = 0.718) or
radiation therapy (uncorrected P-value = 0.431).

@@

\textbf{Figure 3: Graph of the Receiver Operating Characteristic Curve
on Test Set Performance of the Best, Middle, Worse, and Full Training
Models.} For each of the 100 training cohort sets used had 392
individuals and the testing cohort sets had 13 individuals. The AUC on
the test sets for the best, middle, and worse models from training were
0.646, 0.744, and 0.904, respectively. cvAUC is the 20 times repeated
10-fold cross-validated AUC from training.
